---
title: "Case Study"
author: "Brody Erlandson"
output: pdf_document
---

# Introduction

In this vignette, we walk through the code used to run the case study in the 
paper "A Bayesian Functional Concurrent Zero-Inflated Dirichlet Multinomial 
Regression Model". This also serves as a demonstration of the `FunCZIDM` 
package. To run multiple chains, we used a `slurm` array within a high 
performance computing cluster to sample in parallel. Both the `.R` and 
`slurm` scripts are provided for ease of reproducability and implementation.
TODO : Where are they provided?

# Setup 

The data used in the case study is available in the `FunCZIDM` package.

```{r}
library(FunCZIDM)

data(infantCounts)
data(infantCovariates)
```

Though the data is already in the correct format, you should ensure that your 
data is in the correct format. The `FunCZIDM` function requires the 
factors in covariates to be in the correct format. 

```{r}
str(infantCovariates)
```

The  `FunCZIDM` function will use the reference level as the reference category 
for the variable. If you want to change the reference level, you can use the
`relevel` function. The continuous variables will be standardized automatically, 
and the output of `FunCZIDM` will have the means and standard deviations used 
for standardization. This is useful for the other functions, as you will always 
input the original scale of the continuous variables.

The index of the variable to vary by, `Day.of.life.sample.obtained` in the 
case study, and the id's of the subjects, `Subject`, should be noted, as they will be 
used in the `FunCZIDM` function. The `which` function can be used to find the 
index for ease.

```{r}
idIndex <- which(colnames(infantCovariates) == "Subject")
timeIndex <- which(colnames(infantCovariates) == "Day.of.life.sample.obtained")
```

The counts data should be aligned with the covariates data, meaning the $i^{th}$ 
row of the counts data should correspond to the $i^{th}$ row of the covariates
data. The counts data should be a matrix, and only contain counts (i.e., no id 
column).

```{r}
head(infantCounts)
```

# Running the Model

Once the data is in the correct format, you can sample from the FunC-ZIDM model 
using the `FunCZIDM` function. A full description of the arguments and options 
can be found in the reference manual. 

```{r, eval=FALSE}
# This code is not run evaluated.
FunCZIDM(infantCounts, 
         infantCovariates, 
         idIndex,
         timeIndex,
         iter=...,
         burnIn=...,
         thin=10,
         toReturn=c("RA"),
         priors=list("first beta sd"=1, "a"=3, "b"=3, "alpha"=.01, 
                     "beta"=10),
         saveToFile=TRUE,
         saveNullInBetaCI=TRUE,
         nullInBetaCIFileName="nullInBetaCI.csv")
```

Since we are ran multiple processes, we save the output to a file and combine 
all the chains after the each has finished sampling. The `FunCZIDM` package 
includes `combineOutputs` function to combine these files, allowing for ease of 
processing the results. 

# Inference

The package includes the combined output from the case study, so it is not 
nessicary to run the sampling. We demonstrate how to use the functions to get 
inference from the samples using the preloaded samples. First, we show how one 
would combine the files.

```{r, eval=FALSE}
# This code is not evaluated.
outputFiles <- c("output1.rds", "output2.rds", "output3.rds", "output4.rds")
combinedOutput <- combineOutputs(outputFiles, saveToFile=FALSE)
```

With the combined output, the package has the ability to calculate the relative 
abundance and $\alpha$-diversity functions ($\text{RA}_j(t)$ and $\alpha_p(t)$),
the change in relative abundance and $\alpha$-diversity functions 
($\delta_v \alpha_p(t)$ and $\delta_v \text{RA}_{jp}(t)$), and the functional 
coefficients ($\beta_{jp}(t)$). Note in the case study script we do not 
calculate any of these statistics. Instead, we check the `nullInBetaCI.csv` file
to see which covariates may have significant 
